# Introduction

## Dataset

The dataset chosen for the project is
the Barcelona metro passengers (thousands of passengers). Monthly Data from
1996 to 2019 [@noauthor_transporte_nodate].
Figure \@ref(fig:000-plot) shows the plot of the time series.

```{r 000-data}
metro <- window(ts(read.table("./data/metro.dat"), start = 1996, freq = 12))
```

The dataset has information on the number of transported passengers on the
Barcelona metropolitan area metro network. In the last few years,
the government of the city of Barcelona has been pushing towards a
reduction of cars in the city. The aim is to reduce contamination and
pollution as well as make the city safer for pedestrians
\cite{324cat_reduir_2022,noauthor_pla_nodate,noauthor_pla_2018}.

This move towards reducing the use of private transport in Barcelona
has increased the use of public transport.

```{r 000-plot, echo=FALSE, fig.width=6, dev="tikz", fig.cap="Barcelona monthly metro passengers"}
library(tidyverse)
ggplot(data.frame(metro), aes(x = time(metro), y = metro)) +
    geom_line() +
    labs(title = "Barcelona metro passengers", y = "Thousands of Passengers") +
    # theme_bw() +
    scale_x_continuous(minor_breaks = seq(1996, 2020, 1)) +
    scale_y_continuous(limits = c(0, 40000), expand = c(0, 0)) +
    theme(
          panel.grid.minor.x = element_line(colour = 4, linetype = 2),
          panel.grid.major.x = element_line(colour = 4, linetype = 2)
          ) +
    labs(x = "Year")
```
