# Identification

```{r 01-libraries, include=FALSE}
library(tidyverse)
library(forecast)
library(patchwork)
```

## Stationarity

### Constant variance

In order to check for non-constant variance in the time series, we can use
a mean-variance plot. To do so, we group the data in groups of 12
months and calculate the mean and variance of each group.

```{r 010-mean-variance, fig.cap="Mean-variance plot"}
metro <- window(ts(read.table("./data/metro.dat"), start = 1996, freq = 12))

df <- tibble(passengers = metro, date = time(metro))
df %>%
    group_by(year = floor(date)) %>%
    summarise(mean = mean(passengers), var = var(passengers)) %>%
    ggplot(aes(x = mean, y = var)) +
    geom_point() +
    geom_path(alpha = 0.1) +
    geom_text(aes(label = year),
        vjust = -0.5, hjust = 0.5,
        position = position_dodge(width = .9)
    ) +
    labs(x = "Mean", y = "Variance")
```

With the mean variance plot of \@ref(fig:010-mean-variance), there is
no apparent trend in the variance of the data. This suggests that the
time series is stationary. The visualization of \@ref(fig:010-boxplot)
also shows that the variance is more or less sustained over time, except
a few outliers. Therefore, the variance is constant and there is
no need to apply a transformation to the data.

```{r 010-boxplot, fig.cap="Boxplot of the data grouped by year"}
df %>%
    mutate(year = floor(date)) %>%
    ggplot(aes(group = year, x = year, y = passengers)) +
    geom_boxplot() +
    labs(x = "Year", y = "Thousands of Passengers")
```

### Seasonality

Given that the data comes from a monthly time series, and that
the use of public transport is probably affected by the season,
we check some plots to see if that is the case:

```{r 010-decomp, fig.cap="Decomposition of the time series"}
metro %>%
    decompose(filter = rep(1 / 12, 12)) %>%
    autoplot() & scale_x_continuous(minor_breaks = seq(1996, 2020, 1))
```

```{r 010-seasonal-difference, fig.cap="Seasonal difference"}
ggmonthplot(metro) + labs(y = "Thousands of Passengers")
```

```{r 010-seasonplot, fig.cap="Seasonal plot"}
ggseasonplot(metro, year.labels = TRUE)
```

Doing a `monthplot` as shown in \@ref(fig:010-seasonal-difference),
we can see that the data has a yearly seasonality. If we decompose
the series, we obtain the plot shown in \@ref(fig:010-decomp).

We can observe a clear seasonal pattern, people travel less in august,
where most people have holidays, and thus they use less public transport.

We eliminate the seasonality by applying a seasonal difference
$(1 -B^{12}) X_t$.

```{r 010-seasonal-diff, echo=TRUE}
d12metro <- diff(metro, lag = 12)
```

### Constant mean

Once we have removed the seasonality, we can check for a constant
mean.

```{r 010-regular-diff, echo=TRUE}
d1d12metro <- diff(d12metro, lag = 1)
d1d1d12metro <- diff(d1d12metro, lag = 1)
```

```{r 010-time-series, fig.cap="Transformation into stationary series"}
pd12 <- ggplot(NULL, aes(x = time(d12metro), y = d12metro)) +
    geom_line() +
    geom_hline(yintercept = mean(d12metro), linewidth = 1, color = 2) +
    labs(
        title = "$(1-B^{12})X_t$",
        subtitle = paste0("Variance: ", round(var(d12metro))),
        y = "\\texttt{d12metro}"
    )

pd1d12 <- ggplot(NULL, aes(x = time(d1d12metro), y = d1d12metro)) +
    geom_line() +
    geom_hline(yintercept = mean(d1d12metro), linewidth = 1, color = 2) +
    labs(
        title = "$(1-B)(1-B^{12})X_t$",
        subtitle = paste0("Variance: ", round(var(d1d12metro))),
        y = "\\texttt{d1d12metro}"
    )

# pd1d1d12 <- ggplot(NULL, aes(x = time(d1d1d12metro), y = d1d1d12metro)) +
#     geom_line() +
#     geom_hline(yintercept = mean(d1d1d12metro), linewidth = 1, color = 2) +
#     labs(
#         title = "$(1-B)(1-B)(1-B^{12})X_t$",
#         subtitle = paste0("Variance: ", round(var(d1d1d12metro))),
#         y = "\\texttt{d1d1d12metro}"
#     )

pd12 + pd1d12 &
    labs(x = "Year") &
    scale_x_continuous(minor_breaks = seq(1996, 2020, 1))
```

In figure \@ref(fig:010-time-series), we can see that without
applying a regular difference, we do not obtain a constant mean.
With one regular difference, the mean seems constant, 
if we check the variance, for a second regular difference, the variance
increases, indicating that a second regular difference is **not** needed:

```{r 010-variance, fig.width=3, fig.cap="Variance values"}
tibble(
    diff = c("0", "1", "2"),
    var = c(var(d12metro), var(d1d12metro), var(d1d1d12metro))
) %>%
    ggplot(aes(x = diff, y = var)) +
    geom_col() +
    geom_text(aes(label = round(var)), vjust = -0.5, hjust = 0.5) +
    labs(x = "Regular differences", y = "Variance")
```

Moreover, looking at the variance values, the variance also increases with
the first regular difference, which would indicate that we are
over-differencing the data and that we should not apply a regular difference.

However, \@ref(fig:010-time-series) does not seem to display a constant mean,
so we consider the first regular difference as an adequate transformation.

Thus, our proposed transformation is:
\begin{equation*}
W_t = (1-B)(1-B^{12})X_t
\end{equation*}
and the resulting double differenced series (1 seasonal with $s=12$ and 1 regular) `d1d12metro` is considered stationary with seemingly
zero mean. We have that: $d=1$, $D=1$ and $s=12$.


## ACF and PACF

Let's now check the autocorrelation and partial autocorrelation of the
time series. An analyze them in order
to choose the adequate ARIMA model.

First, we are gonna check the ACF and PACF of `d12metro`, that is,
the series without the regular difference.

#### Without regular difference

```{r 010-acf-pacf-non-regular, echo=FALSE, dev="tikz", fig.width=7, fig.cap="ACF and PACF of series without regular difference"}
ggAcf(d12metro, lag.max = 48) + plot_spacer() + ggPacf(d12metro, lag.max = 48) +
    plot_annotation(title = "Series: \\texttt{d12metro}") +
    plot_layout(widths = c(4, 0.5, 4)) &
    ylim(-1, 1) & labs(title = "") &
    theme(aspect.ratio = 1) &
    scale_x_continuous(
        breaks = seq(0, 48, 12),
        # minor_breaks = seq(3, 24, 3)
    )
```

On the PACF plot, we can see strong correlation at lag 1, 2, 3 and a decreasing
pattern in the ACF, which suggests an $AR(3)$ model. For the seasonal
components, we see a strong correlation in the ACF at lag 12,
and a decreasing pattern in the PACF,
which suggests an $MA(1)$ model. Thus, we propose the following
SARIMA model: $SARIMA(3,0,0)(0,0,1)_{12}$.

If we now check the model without the regular difference and test the T-ratios
of the coefficients, we find that the intercept is statistically significant,
therefore, the data is not stationary, as suspected, we need to apply a
the regular difference.

```{r 010-model, echo=TRUE}
(mod <- arima(d12metro, order = c(3, 0, 0), seasonal = list(order = c(0, 0, 1), period = 12)))
abs(mod$coef / sqrt(diag(mod$var.coef)))
```

#### With regular difference

Now, we consider the series `d1d12metro`, that is, the series with the
regular difference.

```{r 010-acf-pacf-regular, echo=FALSE, dev="tikz", fig.width=7, fig.cap="ACF and PACF of the series with regular difference"}
ggAcf(d1d12metro, lag.max = 48) + plot_spacer() + ggPacf(d1d12metro, lag.max = 48) +
    plot_annotation(title = "Series: \\texttt{d1d12metro}") +
    plot_layout(widths = c(4, 0.5, 4)) &
    ylim(-1, 1) & labs(title = "") &
    theme(aspect.ratio = 1) &
    scale_x_continuous(
        breaks = seq(0, 48, 12),
        # minor_breaks = seq(3, 24, 3)
    )
```

This ACF-PACF plot is not as easy to interpret as the previous one, we can
interpret at least these two candidate models:

- $AR(2)$: since, we can see a strong correlation at lags 1 and 2 of the PACF
  and a decreasing pattern in the ACF.
- $MA(1)$: we can see a strong correlation at lag 1 in the ACF and a decreasing
  pattern in the PACF.

For the seasonal component, we see a strong correlation at lag 1 and 12 of the
ACF and nothing relevant in the PACF, so we opt for a $SMA(2)$.

Our models are: $SARIMA(2,0,0)(0,0,2)_{12}$ and $SARIMA(0,0,1)(0,0,2)_{12}$ on
the transformed data series `d1d12metro`.

```{r 010-models, echo=TRUE}
(mod1 <- arima(d1d12metro, order = c(2, 0, 0), seasonal = list(order = c(0, 0, 2), period = 12)))
(mod2 <- arima(d1d12metro, order = c(0, 0, 1), seasonal = list(order = c(0, 0, 2), period = 12)))

abs(mod1$coef / sqrt(diag(mod1$var.coef)))
abs(mod2$coef / sqrt(diag(mod2$var.coef)))
```


